---
title: "directories_mgmt"
author: "geodevm"
date: "2/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library(here)
library(readr)
library(tidyr)
library(tibble)
```

## Using here, readr, tibble, and tidyr package to manage directories and data import/export.

By using this package and sending self-contained projects to collaborators, I can avoid the issues that come with needing to set new working directories upon reciept.

You can use various read-in options from the readr package to efficiently read in different data types. Package readr is also imported via the tidyverse package.

```{r}
# Comma Delimited
data_frame <- read_csv(here("data", "data_frame.csv"))
# Semi-colon Delimited: read_csv2()
# Generalized Delimiter: read_delim(, delim = "")
# Fixed Width Files: read_fwf()
# Tab Delimited Files: read_tsv() or read_table()
# Read Single String: read_file(, locale = default_locale())
```

You can also use some of these arguments to make imports more specific.

```{r}
# No header
read_csv(here("data", "data_frame.csv"), col_names = FALSE)
# Specified header
read_csv(here("data", "data_frame.csv"), col_names = c("X", "Y"))
# Skip lines
read_csv(here("data", "data_frame.csv"), skip = 1)
# Subset
read_csv(here("data", "data_frame.csv"), n_max = 3)
# Specify NAs
read_csv(here("data", "data_frame.csv"), na = "D")
# Parse columns explicitly rather than default col_guess()
read_csv(here("data", "data_frame.csv"), col_types = cols(letter = col_factor()))
# Can also use:
# col_character()
# col_factor(levels, ordered = FALSE)
# col_double()
# col_datetime(format = "")
# col_date(format = "")
# col_time(format = "")
# col_integer()
# col_logical()
# col_number() or col_numeric()
# col_skip()
# They can also be parsed after the fact with the same endings, with prefix parse (e.g. parse_factor())
```

Use readr::problems to diagnose any problems with reading in a dataset.

```{r}
problems(data_frame)
```
Converting to tibbles for work in R.

```{r}
data_tib <- as_tibble(data_frame)
is_tibble(data_tib)
```

Subsetting tibbles:

```{r}
# Return a tibble:
data_tib[, "letter"]
# Return a vector
data_tib[["letter"]]
data_tib$letter
```

View tibbles with glimpse()

```{r}
glimpse(data_tib)
# Or open in its own window:
View(data_tib)
```

Reshape data with tidyr::gather() and tidyr::spread()

```{r}
# Create a non-tidy tibble to be gathered, with a column specified rather than row per observation:
gather_tib <- tibble(letter = c("a", "b"), "1900" = 1:2, "2000" = 4:5)
gather(gather_tib, "1900", "2000", key = "year", value = "number")
# Create a non-tidy tibble to be spread, with multiple column values in a column:
spread_tib <- tibble(letter = c("a", "a", "a", "a", "b", "b", "b", "b"),
                     year = c("1900", "1900", "2000", "2000", "1900", "1900", "2000", "2000"), 
                     type = c("c", "d", "c", "d", "c", "d", "c", "d"),
                     count = 1:8)
spread(spread_tib, type, count)
```

Manage NA values

```{r}
# Create a tibble with NA values
na_tib <- tibble(X = c("a", "b", "c", "d"), Y = c(1, NA, 3, NA))
# Drop rows with NA values
drop_na(na_tib, Y)
# Fill in missing values with most recent non-NA
fill(na_tib, Y)
# Replace NAs with a certain value
replace_na(na_tib, list(Y = 2))
```

Split cells into multiple columns or combine them.

```{r}
# Create a tibble:
sep_tib <- tibble(letter = c("a", "b", "c", "d"), year = c("19/00", "19/10", "19/20", "19/30"))
# Use separate to separate the year into two parts
unite_tib <- separate(sep_tib, year, sep = "/", into = c("century", "year"))
unite_tib
# Combine those back together with unite
unite(unite_tib, century, year, col = "year", sep = "")
```


readr also offers a variety of options for exporting data in different formats.

```{r}
# Comma Delimited
write_csv(data_frame, "data", na = "NA", append = FALSE)
# Generalized Delimiter: write_delim(, delim = "")
# String to File: write_file(data_frame, "data", append = FALSE)
# Excel csv: write_excel_csv()
# String Vector to File, Line/Element: write_lines()
```

Session:

```{r}
sessionInfo()
```