---
title: "tidyverse_flows"
author: "geodevm"
date: "2/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(nycflights13)
```

##Data visualization.

We'll be using the mpg dataset from ggplot2.

```{r}
mpg
```

Now we'll create a gglplot using geom_point.


```{r}
ggplot(mpg) +
  geom_point(aes(displ, hwy))
# The template for these graphs would be the following:
# ggplot(data = <DATA>) +
#   <GEOM_FN>(mapping = aes(<MAPPINGS: x = x, y = y>))
```

Create aesthetic mappings.

```{r}
ggplot(mpg) +
  geom_point(aes(displ, hwy, color = class))
# Can also use:
#   size =
ggplot(mpg) +
  geom_point(aes(displ, hwy, size = class))
#   alpha =
ggplot(mpg) +
  geom_point(aes(displ, hwy, alpha = class))
#   shape =
ggplot(mpg) +
  geom_point(aes(displ, hwy, shape = class))
```

We can also just define the aesthetics outside of the mapping argument.

```{r}
ggplot(mpg) +
  geom_point(mapping = aes(displ, hwy), color = "blue")
```

This .png shows the different types of shapes that can be used.

https://d33wubrfki0l68.cloudfront.net/e28a1b57b6622cf67fd8a7e01c6a9955914f8fe9/635be/visualize_files/figure-html/shapes-1.png

Size can be specified in mm.

To facet within one dataset, use the facet_wrap() function

```{r}
ggplot(mpg) +
  geom_point(aes(displ, hwy)) +
  facet_wrap(~ class, nrow = 2)
```

The facet_grid() function is another way to facet on two-variable axes.

```{r}
ggplot(mpg) +
  geom_point(aes(displ, hwy)) +
  facet_grid(drv ~ cyl)
```

If you only want to display rows or columns, using "." as an argument for facet grid will ensure that there is only one dimension.

```{r}
ggplot(mpg) +
  geom_point(aes(displ, hwy)) +
  facet_grid(. ~ cyl)
ggplot(mpg) +
  geom_point(aes(displ, hwy)) +
  facet_grid(drv ~ .)
```

Now for a few examples of different geoms

```{r}
ggplot(mpg) + 
  geom_smooth(aes(displ, hwy))
ggplot(mpg) + 
  geom_smooth(aes(displ, hwy, linetype = drv))
ggplot(mpg, aes(displ, hwy)) + 
  geom_smooth(aes(linetype = drv, color = drv)) +
  geom_point(aes(color = drv))
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth()
```

Mixed mapping aesthetics:

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = class)) + 
  geom_smooth()
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

Stat Transformations

geom_bar uses a count transformation

```{r}
ggplot(diamonds) +
  geom_bar(aes(cut))
# Same as:
ggplot(diamonds) + 
  stat_count(aes(cut))
```

You can change the stat of certain type of plot:

```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)

ggplot(demo) +
  geom_bar(aes(cut, freq), stat = "identity")
```

You can also do things like display the proportion instead of the count:

```{r}
ggplot(diamonds) + 
  geom_bar(aes(cut, stat(prop), group = 1))
```

Or you can highlight different values within the data set with a summary plot:

```{r}
ggplot(diamonds) + 
  stat_summary(aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

Bar chart adjustments:

```{r}
ggplot(diamonds) + 
  geom_bar(aes(cut, color = cut))
ggplot(diamonds) + 
  geom_bar(aes(cut, fill = cut))
ggplot(diamonds) + 
  geom_bar(aes(cut, fill = clarity))
ggplot(diamonds, aes(cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
ggplot(diamonds, aes(cut, color = clarity)) + 
  geom_bar(fill = NA, position = "identity")
ggplot(diamonds) + 
  geom_bar(aes(cut, fill = clarity), position = "fill")
ggplot(diamonds) + 
  geom_bar(aes(cut, fill = clarity), position = "dodge")
```

Enhanced visualization of scatter plots using jitter.

```{r}
ggplot(mpg) + 
  geom_point(aes(displ, hwy))
ggplot(mpg) + 
  geom_point(aes(displ, hwy), position = "jitter")
```

Coordinate flips can be useful for graphs where a categorical variable might appear on the x axis:

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```

If you're doing a map, you can also adjust the coordinates in ggplot to transform into a more readable geographic transformation

```{r}
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()
```

If we want the coordinates to flip to become polar, we can use coord_polar()

```{r}
bar <- ggplot(diamonds) + 
  geom_bar(aes(cut, fill = cut), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

The overall structure of ggplot is:

ggplot(<DATA>) + 
  <GEOM_FUNCTION(S)>(
     aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
  
## Data Transformation.

```{r}
flights
```

Some of the data types in tibbles:
int: integer
dbl: double (real numbers)
chr: character/string
dttm: date-times
lgl: logical, T/F
fctr: factors
date: date

Main dplyr functions:
filter() picks observations based on values
arrange() reorders rows
select() picks variables by their names
mutate() creates a new variables with functions of existing variables
summarise() collapses many values into a single summary
group_by() changes the scope of the above functions

### Using filter():

```{r}
jan1 <- filter(flights, month == 1, day == 1)
jan1

dec25 <- filter(flights, month == 12, day == 25)
dec25
```

Package dplyr also offers near() to evaluate floats, where there may be slight imprecision in testing equality. This could be useful for filtering.

```{r}
sqrt(2) ^ 2 == 2
1 / 49 * 49 == 1
near(sqrt(2) ^ 2, 2)
near(1 / 49 * 49, 1)
```

The Boolean operators for filter are...
&: "and"
|: "or"
!: "not"

Don't use && or || for filter arguments.

Other useful operators...
%in%: x %in% y selects every row where x is one of the values in y.

There are usually multiple ways to write the same thing:

```{r}
filter(flights, month == 11| month == 12)
filter(flights, month %in% c(11, 12))
```

```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```

Filter excludes rows that are NA by default. IF you want to preserve that, is.na must be specified as a condition. 

```{r}
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)
filter(df, is.na(x) | x > 1)
```

### Using arrange()

This function seems to be somewhat limited in scope. Just reorders rows.

```{r}
arrange(flights, year, month, day)
```

Re-order with desc() for descending.

```{r}
arrange(flights, desc(dep_delay))
```

NAs are always at the end of the list in the column.

### Using select()

This function is used to select specified columns, essentially subsetting by column.

```{r}
select(flights, year, month, day)
# Specify a series of columns to include:
select(flights, year:day)
# Specify a series of columns to exclude:
select(flights, -(year:day))
```

Within select, there are a number of specifying functions.
starts_with(): select columns starting with a certain string
ends_with(): select columns ending with a certain string
contains(): select columns that contain a certain string
matches(): select variables that match a regular expression within the string
num_range(): example: num_range("x", 1:3) matches x1, x2, x3

Rename is a version of select() that will select a column and rename it while conserving its properties within the tibble.

```{r}
rename(flights, tail_num = tailnum)
```

everything() can also be used to conserve columns.

```{r}
select(flights, time_hour, air_time, everything())
#This reorders the time and air time columns while conserving all the columns afterward.
```

### Using mutate()

Mutate can be used to calculate new columns

```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
# If we want to use columns created within the mutate call to create a new calculated column, that can be done with order of operations.
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
# And if you only want to keep the new variables, use transmutate.
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

Useful functions to use within mutate:
+
-
*
/
^
mean(): example x / mean(x)
sum()
other aggregate functions
%/% (integer division)
%% (remainder)
example:
```{r}
transmute(flights,
 dep_time,
 hour = dep_time %/% 100,
 minute = dep_time %% 100
)
```
log()
log2()
log10()
lead()
lag()
example:
```{r}
x <- 1:10
lag(x)
lead(x)
```
cumsum() (cumulative)
cummean()
example:
```{r}
cumsum(x)
cummean(x)
```
<
>
<=
>=
!=
==
min_rank()
row_number()
dense_rank()
percent_rank()
cume_dist()
